---
title: "SOM3 - Reading and ploting data"
author: "Paixao et al. 2021. The Middle Paleolithic Ground Stones Tools of Nesher Ramla Unit V (Southern Levant): a multi-scale use-wear approach for assessing the assemblage functional variability. Quaternary International"
date: "`r Sys.time()`"
output:
  html_document: default
  word_document: default
  pdf_document: default
geometry: margin=1in
fontsize: 9pt
---

\pagenumbering{gobble}

```{r Knitr Options, include=F}
knitr::opts_chunk$set(comment = NA, message = FALSE, indent = "", error = TRUE)

```

__Brief description of the script__

This R markdown document reads, summarizes and plots data for: *Paixao et al. 2021. The Middle Paleolithic Ground Stones Tools of Nesher Ramla Unit V (Southern Levant): a multi-scale use-wear approach for assessing the assemblage functional variability. Quaternary International*. The document contains:

1. Manuscript's tables, 
2. Manuscript's figures (data plots),
3. Supplementary material, including extra tables and figures (data plots)

This R project and respective script follows the procedures described by Marwick et al. 2017. To compile this markdown document do not delete or move files from their original folders.
Please note that the tables and figures numbering in this file do not match the numbering in the original manuscript.

For any questions, comments and inputs, please contact:

Eduardo Paixão. paixao@rgzm.de, or João Marreiros. marreiros@rgzm.de

# Load data into R project

*The imported files are in: `r paste0("../data/raw_data")`*

*All figures will be saved in: ´r past0("../figures")'*

*All tables will be saved in: ´r past0("../data/derived_data")'*


```{r}

# Load required libraries

library(tidyverse)
library(utils)
library(knitr)
library(janitor)
library(flextable)
library(GGally)
library(doBy)
library(ggpubr)
library(tools)

# See your WD and update the following paths
# getwd()

# Load data from .csv
db1 <- read.csv("../raw_data/generaldb.csv", header=T, ",") # general database
db2 <- read.csv("../raw_data/functionaldb.csv", header=T, ",") # dataset related to functional 

# On db1, convert strings to numeric and replace "," by "."
db1$length <- as.numeric(gsub(",", ".", db1$length))
db1$width <- as.numeric(gsub(",", ".", db1$width))
db1$thickness <- as.numeric(gsub(",", ".", db1$thickness))
db1$weight <- as.numeric(gsub(",", ".", db1$weight))

# Filter general dataset (only Nesher Ramla is selected here)
db1nr <- filter(db1, site  == "Nesher Ramla")

data_file <- list.files("../analysis/raw_data/", pattern = "\\.csv$", full.names = TRUE)

```

\newpage


# Tables
## General Inventory
```{r results = "asis"}
# Nesher Ramla

inventory <- db1nr %>%
  group_by(raw_material, support) %>%
  summarize(total = n()) %>%
  pivot_wider(names_from = "support",
              values_from = "total",
              values_fill = 0) %>%
  adorn_totals(where = c("row", "col"), fill = "") %>%
  rename("Raw material" = raw_material, Boulder = boulder, Block = block)
write_csv(inventory, "../derived_data/generalinventory_nesher.csv")

table <- flextable(inventory)
table <- set_caption(table,caption = "General inventory of Grounds Stone tools 
            from Nesher Layer 5")
knit_print(table)

```

## Typological inventory
```{r results = "asis"}

# Nesher Ramla
type <- db1nr %>%
  group_by(raw_material, typology) %>%
  summarize(total = n()) %>%
  pivot_wider(names_from = "typology",
              values_from = "total",
              values_fill = 0) %>%
  adorn_totals(where = c("row", "col"), fill = "") %>%
  rename("Raw material" = raw_material)
write_csv(type, "../derived_data/typeinventory_nesher.csv")

table <- flextable(type)
table <- set_caption(table,caption = "Typological inventory by raw material from Nesher Ramla, Layer 5")
knit_print(table)

```

## Morphometric analysis
```{r, fig.height = 5.5, fig.width = 8.5, res = 300}

# Nesher Ramla

ggplot(db1nr, aes (x = length, y = width, color = support)) + 
  geom_point(size=0.5) +
  stat_ellipse() +
  labs(x="Length (mm)", y="Width (mm)", color = "Support") +
  facet_wrap(vars(typology)) +
  scale_color_discrete(labels = c("Block", "Boulder", "Pebble")) +
  theme(axis.text.x = element_text(face= "bold"), axis.text.y = element_text(face = "bold"), axis.line = element_line(color = "black", size = 0.5, linetype = "solid"))

ggsave("../plots/metrics_nr.png")

ggplot(db1nr, aes(x = support, y = weight)) + 
  geom_boxplot() + 
  labs(x="Typology", y="Weight (g)", face = "bold") +
  facet_wrap(vars(typology)) +
    scale_x_discrete(labels = c('Block','Boulder','Pebble')) +
  theme(axis.text.x = element_text(face= "bold"), axis.text.y = element_text(face = "bold"), axis.line = element_line(color = "black", size = 0.5, linetype = "solid"))


ggsave("../plots/weight_nr.png")

```
\newpage

## Use-wear macro damage, first observations
```{r results = "asis"}

# Nesher Ramla

# sampling observation with macro wear traces
damagenr <- filter(db1nr, usewear_macro == 'yes')

# Use-wear damage by tool typology 

macrodamage <- damagenr %>%
  group_by(typology, usewear_type) %>%
  summarize(total = n()) %>%
  pivot_wider(names_from = "usewear_type",
              values_from = "total",
              values_fill = 0) %>%
  adorn_totals(where = c("row", "col"), fill = "") %>%
  rename("Typology" = typology, Mixed = mix, "Impact marks" = pecking, 
         Polish = polish, Striations = striations)
write_csv(macrodamage, "../derived_data/macrodamage_nr.csv")

table <- flextable(macrodamage)
table <- set_caption(table,caption = "Use-wear damage 
            organized by typology, Nesher Ramla - Layer 5")
knit_print(table)

```

## Macro wear analysis
```{r results = "asis"}

# Nesher Ramla

# Macro wear traces by tool typology 

macrowear <- db2nr %>%
  filter(!is.na(MACROTYPE)) %>%
  group_by(TYPOLOGY, MACROTYPE) %>%
  summarize(total = n()) %>%
  pivot_wider(names_from = "MACROTYPE",
              values_from = "total",
              values_fill = 0) %>%
  adorn_totals(where = c("row", "col"), fill = "") %>%
  rename("Typology" = TYPOLOGY, "Type 1" = "1", "Type 2" = "2")
write_csv(macrowear, "../derived_data/macrowear_nr.csv")

table <- flextable(macrowear)
table <- set_caption(table,caption = "Macro wear traces 
            organized by typology, Nesher Ramla - Layer 5")
knit_print(table)

```

## Micro wear analysis
```{r results = "asis"}

# Using db2 dataset, which contains all micro wear observations
# This only applies to Nesher Ramla (see Methods and Results chapter for more info)

microwear <- db2nr %>%
  filter(MICROTYPE %in% c("A", "B","C","D")) %>%
  group_by(TYPOLOGY, MICROTYPE) %>%
  summarize(total = n()) %>%
  pivot_wider(names_from = "MICROTYPE",
              values_from = "total",
              values_fill = 0) %>%
  adorn_totals(where = c("row", "col"), fill = "")%>%
  rename("Typology" = TYPOLOGY)
write_csv(microwear, "../derived_data/tab6.csv")

table <- flextable(microwear)
table <- set_caption(table,caption = "Micro wear type 
            organized by typology, Nesher Ramla - Layer 5")
knit_print(table)

```

## Correlation between macro and micro wear traces
```{r results = "asis"}

# sampling observation with macro wear traces type 1
# This only applies to Nesher Ramla (see Methods and Results chapter for more info)

macro1 <- db2nr %>%
  filter(MACROTYPE == "1", MICROTYPE %in% c("A", "B","C","D")) %>%
  group_by(TYPOLOGY, MICROTYPE) %>%
  summarize(total = n()) %>%
  pivot_wider(names_from = "MICROTYPE",
              values_from = "total",
              values_fill = 0) %>%
  rename("Typology" = TYPOLOGY) %>%
  adorn_totals(where = c("row", "col"), fill = "")
write_csv(microwear, "../derived_data/macro1&micro_nr.csv")

table <- flextable(macro1)
table <- set_caption(table,caption = "Micro wear traces associated with Macro traces type 1, and organized by typology, Nesher Ramla - Layer 5")
knit_print(table)

# sampling observation with macro wear traces type 2
# This only applies to Nesher Ramla (see Methods and Results chapter for more info)

macro2 <- db2nr %>%
  filter(MACROTYPE == "2", MICROTYPE %in% c("A", "B","C","D")) %>%
  group_by(TYPOLOGY, MICROTYPE) %>%
  summarize(total = n()) %>%
  pivot_wider(names_from = "MICROTYPE",
              values_from = "total",
              values_fill = 0) %>%
  rename("Typology" = TYPOLOGY) %>%
  adorn_totals(where = c("row", "col"), fill = "")
write_csv(microwear, "../derived_data/macro2&micro_nr.csv")

table <- flextable(macro2)
table <- set_caption(table,caption = "Micro wear traces associated with Macro traces type 2, and organized by typology, Nesher Ramla - Layer 5")
knit_print(table)

```

## Relation between different active areas and macro wear traces within and between tools
```{r, results = "asis"}

# Nesher Ramla
# 2 impact areas

a <- filter(db2nr, WEAR2 == "yes")
a <- select(a, TYPOLOGY, WEAR1_MARKS_TYPE, WEAR2_MARKS_TYPE)
b <- a %>%
  unite(a, WEAR1_MARKS_TYPE, WEAR2_MARKS_TYPE, sep = "/")

amostra <- b %>%
 count(TYPOLOGY, a)

sampled <- spread(amostra, key = a, value = n)
sampled[is.na(sampled)] <- 0

twoimpac <- sampled %>%
  adorn_totals(where = c("row", "col"), fill = "")

table <- flextable(twoimpac)
table <- set_caption(table,caption = "Tools with 2 active areas 
            with macro wear traces, Nesher Ramla - Layer 5")
knit_print(table)

# 3 impact areas

a <- filter(db2nr, WEAR3 == "yes")
a <- select(a, TYPOLOGY, WEAR1_MARKS_TYPE, WEAR2_MARKS_TYPE, WEAR3_MARKS_TYPE)
b <- a %>%
  unite(a, WEAR1_MARKS_TYPE, WEAR2_MARKS_TYPE, WEAR3_MARKS_TYPE, sep = "/")

amostra <- b %>%
 count(TYPOLOGY, a)

sampled <- spread(amostra, key = a, value = n)
sampled[is.na(sampled)] <- 0

threeimpac <- sampled %>%
  adorn_totals(where = c("row", "col"), fill = "")

table <- flextable(threeimpac)
table <- set_caption(table,caption = "Tools with 3 active areas 
            with macro wear traces, Nesher Ramla - Layer 5")
knit_print(table)

# 4 impact areas

a <- filter(db2nr, WEAR4_ == "yes")
a <- select(a, TYPOLOGY, WEAR1_MARKS_TYPE, WEAR2_MARKS_TYPE,
            WEAR3_MARKS_TYPE, WEAR4_MARKS_TYPE)
b <- a %>%
  unite(a, WEAR1_MARKS_TYPE, WEAR2_MARKS_TYPE, WEAR3_MARKS_TYPE,
        WEAR4_MARKS_TYPE, sep = "/")

amostra <- b %>%
 count(TYPOLOGY, a)

sampled <- spread(amostra, key = a, value = n)
sampled[is.na(sampled)] <- 0

fourimpac <- sampled %>%
  adorn_totals(where = c("row", "col"), fill = "")


table <- flextable(fourimpac)
table <- set_caption(table,caption = "Tools with 4 active areas with 
            macro wear traces, Nesher Ramla - Layer 5")
knit_print(table)

# Use-wear location, files for GIS spatial analysis

# Hammerstones

a <- filter(db2nr, TYPOLOGY == "Hammerstone")

a <- select(a, WEAR1_LOCATION, WEAR1_MARKS_TYPE, WEAR2_MARKS_TYPE, WEAR2_LOCATION,
            WEAR3_MARKS_TYPE, WEAR3_LOCATION,
            WEAR4_MARKS_TYPE, WEAR4_LOCATION)
b <- a %>%
  unite(a, WEAR1_MARKS_TYPE, WEAR2_MARKS_TYPE, WEAR3_MARKS_TYPE,
        WEAR4_MARKS_TYPE, sep = "/")

amostra <- b %>%
 count(WEAR1_LOCATION, WEAR2_LOCATION,
       WEAR3_LOCATION, WEAR4_LOCATION, a)

sampled <- spread(amostra, key = a, value = n)
sampled[is.na(sampled)] <- 0

twoimpac <- sampled %>%
  rename("1st area" = WEAR1_LOCATION, "2nd area" = WEAR2_LOCATION, "3rd area" = WEAR3_LOCATION,
         "4th area" = WEAR4_LOCATION) %>%
  adorn_totals(where = c("row", "col"), fill = "")

table <- flextable(twoimpac)
table <- set_caption(table,caption = "Location of use-wear traces on Hammerstones, Nesher Ramla")
knit_print(table)

# Choppers
a <- filter(db2nr, TYPOLOGY == "Chopper")

a <- select(a, WEAR1_LOCATION, WEAR1_MARKS_TYPE, WEAR2_MARKS_TYPE, WEAR2_LOCATION)
b <- a %>%
  unite(a, WEAR1_MARKS_TYPE, WEAR2_MARKS_TYPE, sep = "/")

amostra <- b %>%
 count(WEAR1_LOCATION, WEAR2_LOCATION,a)

sampled <- spread(amostra, key = a, value = n)
sampled[is.na(sampled)] <- 0

twoimpac <- sampled %>%
  rename("1st area" = WEAR1_LOCATION, "2nd area" = WEAR2_LOCATION) %>%
  adorn_totals(where = c("row", "col"), fill = "")

table <- flextable(twoimpac)
table <- set_caption(table,caption = "Location of use-wear traces on Choppers, Nesher Ramla")
knit_print(table)

# Anvils
a <- filter(db2nr, TYPOLOGY == "Anvil")

a <- select(a, WEAR1_LOCATION, WEAR1_MARKS_TYPE, WEAR2_MARKS_TYPE, WEAR2_LOCATION,
            WEAR3_MARKS_TYPE, WEAR3_LOCATION)
b <- a %>%
  unite(a, WEAR1_MARKS_TYPE, WEAR2_MARKS_TYPE,
        WEAR3_MARKS_TYPE, sep = "/")

amostra <- b %>%
 count(WEAR1_LOCATION, WEAR2_LOCATION,
       WEAR3_LOCATION, a)

sampled <- spread(amostra, key = a, value = n)
sampled[is.na(sampled)] <- 0

twoimpac <- sampled %>%
  rename("1st area" = WEAR1_LOCATION, "2nd area" = WEAR2_LOCATION, "3rd area" = WEAR3_LOCATION) %>%
  adorn_totals(where = c("row", "col"), fill = "")

table <- flextable(twoimpac)
table <- set_caption(table,caption = "Location of use-wear traces on Anvils, Nesher Ramla")
knit_print(table)

```

# End and Session info
```{r}
sessionInfo()
```


